JavaScript as an Embedded DSL

### 2012-03-21 papers/ecoop2012/article.tex
most minor comments done

Here is the JavaScript code generated for this snippet:
\begin{lstlisting}[language=JavaScript]
function test(x0) {
  var x6 = []
  for(var x1=0;x1<x0;x1++){
    var x4 = []
    for(var x2=0;x2<x0;x2++){
      var x3 = x1 * x2
      x4[x2]=x3
    }
    x6.splice.apply(x6, [x6.length,0].concat(x4))
  }
  return x6
}
\end{lstlisting}

@@

Notice that \code{document} has type \code{Rep[Element]}, and needs to
implement the interface of \code{ElementOps}, so that
\code{document.getElementById("canvas")} is well-typed. We achieve
this using an implicit conversion from \code{Rep[Element]} to
\code{ElementOps}.  At the DSL implementation level, the \code{ElementOps}
returned by this implicit conversion needs to generate an IR node for
each method call, as shown in the walkthrough
in section~\ref{sec:defineDslComponent}. For example,
\code{document.getElementById("canvas")} becomes the IR node {\tt\small
  MethodCall(document, "getElementById", List("canvas"))}. This is a
mechanical transformation, implemented by \code{repProxy} using
reflection to intercept method calls and generate IR nodes based on
the method name and the arguments of the invocation. Note that this
use of reflection is purely at staging time, so there is no overhead
in the generated code.
\begin{lstlisting}
trait DomLift extends Dom with JSProxyBase {
  implicit def repToElementOps(x: Rep[Element]): ElementOps =
    repProxy[Element,ElementOps](x)
  implicit def repToCanvasOps(x: Rep[Canvas]): CanvasOps =
    repProxy[Canvas,CanvasOps](x)
  implicit def repToContextOps(x: Rep[Context]): ContextOps =
    repProxy[Context,ContextOps](x)
}
\end{lstlisting}

@@

\subsection{Casting and Optional Runtime Type Checks}\label{sec:casting}
The need for casting arises in a few contexts. One of them is the
boundary between typed and untyped portions of a
program~\cite{wadler-blame,blame-for-all}. Passing a value from an
untyped portion to a typed one usually requires a cast. Another
situation where casts are needed is interaction with external
services. For example, to process data from an external service such
as Twitter, we cast it to its expected type:
\begin{lstlisting}
type TwitterResponse = Array[JSLiteral {val text: String}]
def fetchTweets(username: Rep[String]) = {
 val raw = ajax.get { ... }
 raw.as[TwitterResponse]
}
\end{lstlisting}
A more complete example is provided in section~\ref{sec:cps}. In this
situation, it is useful to generate runtime checks either as an aid
during development and debugging time or as a security mechanism that
validates data coming from an external source. In the example above,
if runtime checks are enabled, by failing early, we obtain a guarantee
that all data returned from \code{fetchTweets} conforms to type
\code{TwitterResponse} which means that any later access to the
\code{text} field of any element of the data array will never fail, and
always return a string.

@@

\code{DynamicRep} exploits a new Scala feature, based on a
marker trait \code{Dynamic}: An expression whose type \code{T} is a
subtype of \code{Dynamic} is subject to the following rewrites:
\begin{itemize}
\item \code{x.f} rewrites to \code{x.selectDynamic("f")},
\item \code{x.m(a, ..., z)} rewrites to \code{x.applyDynamic("m")(a, ..., z)},
\item \code{x.f = v} rewrites to \code{x.updateDynamic("f")(v)}.
\end{itemize}
These rewrites take place when the type \code{T} doesn't statically
have field \code{f} and method \code{m}.

@@

As for implementation, notice that the type of a \code{new JSLiteral
  \{...\}} expression is not \code{JSLiteral \{...\}} but
\code{Rep[JSLiteral \{...\}]}. This is achieved with support from the
Scala-Virtualized compiler. \code{JSLiteral} inherits from a special marker
trait, which indicates its \code{new} expressions should be
reified. So the \code{new} expression is turned into a method call,
with information about all the field definitions. A complication is
that a field definition might reference another field being defined
(such as \code{b} being defined in terms of \code{a} in the example
above). So each definition is represented by its name and a {\it
  function} which takes a self type for the object literal. These
definition functions are evaluated in an order which allows the self
references to be resolved.

@@

There are a few things worth noting here. First, the return type of \code{foo} method is rewritten to be \code{ControlContext}\footnote{The \code{ControlContext} class implements the continuation monad and is provided by Scala's standard library. \code{ControlContext} is similar to C\#'s \code{Task<T>}, but more general.}. This is due to the fact that \code{sleep} is used in the body of \code{foo}. Also, note that the code to be executed after sleeping is captured as a continuation (anynomous function) and passed to the \code{ControlContext} through a call of the \code{map} method. The body of the \code{reset} block is rewritten in a similar vein.

@@

\subsection{CPS for Interruptible Traversals}
We need to be able to suspend our execution while traversing an array in order to implement functionality from listing~\ref{code:twitter_example}. Let's consider a simplified example where we want to iterate over an array and sleep during each iteration. We present both code written in JavaScript and our DSL that achieves that (listings~\ref{code:js:sleep_iter}~\&~\ref{code:sleep_iter}).
\begin{figure}
\lstset{basicstyle=\ttfamily\scriptsize}
\begin{changemargin}{-0.7cm}{-0.7cm}
\setlength{\columnseprule}{0.5pt}
\begin{multicols}{2}
\begin{lstlisting}[caption=JavaScript,label=code:js:sleep_iter]
var xs = [1, 2, 3]
var i = 0
var msg = null
function f1() {
  if (i < xs.length) {
    window.setTimeout(f2, xs[i]*1000)
    msg = xs[i]
    i++
  } else {
    console.log("done")
  }
}
function f2() {
  console.log(msg)
  f1()
}
f1()
\end{lstlisting}
\begin{lstlisting}[escapechar=|,caption=DSL,label=code:sleep_iter]
val xs = array(1, 2, 3)
// shorthand for xs.suspendable.foreach
for (x <- xs.suspendable) {
  sleep(x * 1000)
  console.log(String.valueOf(x))
}
log("done")

@@

In the DSL code, we use a \code{suspendable} variant of arrays, which is achieved through an implicit conversion from regular arrays:
\begin{lstlisting}
implicit def richArray(xs: Rep[Array[A]]) = new {
  def suspendable: SArrayOps[A] = new SArrayOps[A](xs)
}
\end{lstlisting}
The idea behind \code{suspendable} arrays is that iteration over them can be interrupted. We'll have a closer look at how to achieve that with the help of CPS. The \code{suspendable} method returns a new instance of the \code{SArrayOps} class defined here:

\begin{lstlisting}[caption=Suspendable foreach,label=code:suspendable_foreach]
class SArrayOps(xs: Rep[Array[A]]) {
  def foreach(yld: Rep[A] => Rep[Unit] @suspendable):
    Rep[Unit] @suspendable = {
      var i = 0
      suspendableWhile(i < xs.length) { yld(xs(i)); i += 1 }
    }
}
\end{lstlisting}
Note that one cannot use while loops in CPS but we can simulate them with recursive functions. Let's see how regular while loop can be simulated with a recursive function:
\begin{lstlisting}
def recursiveWhile(cond: => Boolean)(body: => Unit): Unit = {
  def rec = () => if (cond) { body; rec() } else {}
  rec()
}
\end{lstlisting}

@@

With the abstractions for interruptible loops and traversals at hand,
what remains to complete the Twitter example from the beginning of the section
is the actual Ajax request/response cycle.

>>>>

Here is the JavaScript code generated for this snippet:
\begin{lstlisting}[language=JavaScript]
function test(x0) {
  var x6 = []
  for(var x1=0;x1<x0;x1++){
    var x4 = []
    for(var x2=0;x2<x0;x2++){
      var x3 = x1 * x2
      x4[x2]=x3
    }
    x6.splice.apply(x6, [x6.length,0].concat(x4))
  }
  return x6
}

The generated code ressembles single-assignment form. The nested
\code{for}-loop is desugared into a \code{flatMap} which generates the
nested \code{for}-loop and the \code{splice} pattern concatenating the
inner \code{x4} arrays into one \code{x6} array in the JavaScript
code.\footnote{Obviously, the generated code can be optimized further.}

\end{lstlisting}

@@

Notice that \code{document} has type \code{Rep[Element]}, and needs to
implement the interface of \code{ElementOps}, so that
\code{document.getElementById("canvas")} is well-typed. We achieve
this using an implicit conversion from \code{Rep[Element]} to
\code{ElementOps}.  At the DSL implementation level, the \code{ElementOps}
returned by this implicit conversion needs to generate an IR node for
each method call, as shown in the walkthrough
in section~\ref{sec:defineDslComponent}. For example,
\code{document.getElementById("canvas")} becomes the IR node {\tt\small
  MethodCall(document, "getElementById", List("canvas"))}. This is a
mechanical transformation, implemented by \code{repProxy}, a library method using
reflection to intercept method calls and generate IR nodes based on
the method name and the arguments of the invocation. Note that this
use of reflection is purely at staging time, so there is no overhead
in the generated code.
\begin{lstlisting}
trait DomLift extends Dom with JSProxyBase {
  implicit def repToElementOps(x: Rep[Element]): ElementOps =
    repProxy[Element,ElementOps](x)
  implicit def repToCanvasOps(x: Rep[Canvas]): CanvasOps =
    repProxy[Canvas,CanvasOps](x)
  implicit def repToContextOps(x: Rep[Context]): ContextOps =
    repProxy[Context,ContextOps](x)
}
\end{lstlisting}

@@

\subsection{Casting and Optional Runtime Type Checks}\label{sec:casting}
The need for casting arises in a few contexts. One of them is the
boundary between typed and untyped portions of a
program~\cite{wadler-blame,blame-for-all}. Passing a value from an
untyped portion to a typed one usually requires a cast. Another
situation where casts are needed is interaction with external
services. For example, to process data from an external service such
as Twitter, we cast it to its expected type (an array of JSON objects, each with a field called \code{text}):
\begin{lstlisting}
type TwitterResponse = Array[JSLiteral {val text: String}]
def fetchTweets(username: Rep[String]) = {
 val raw = ajax.get { ... }
 raw.as[TwitterResponse]
}
\end{lstlisting}
A more complete example is provided in section~\ref{sec:cps}. In this
situation, it is useful to generate runtime checks either as an aid
during development and debugging time or as a security mechanism that
validates data coming from an external source. In the example above,
if runtime checks are enabled, by failing early, we obtain a guarantee
that all data returned from \code{fetchTweets} conforms to type
\code{TwitterResponse} which means that any later access to the
\code{text} field of any element of the data array will never fail, and
always return a string.

@@

\code{DynamicRep} exploits a new Scala feature, based on a
special trait \code{Dynamic}: An expression whose type \code{T} is a
subtype of \code{Dynamic} is subject to the following rewrites:
\begin{itemize}
\item \code{x.f} rewrites to \code{x.selectDynamic("f")},
\item \code{x.m(a, ..., z)} rewrites to \code{x.applyDynamic("m")(a, ..., z)},
\item \code{x.f = v} rewrites to \code{x.updateDynamic("f")(v)}.
\end{itemize}
These rewrites take place when the type \code{T} doesn't statically
have field \code{f} and method \code{m}.

@@

As for implementation, notice that the type of a \code{new JSLiteral
  \{...\}} expression is not \code{JSLiteral \{...\}} but
\code{Rep[JSLiteral \{...\}]}. This is achieved with support from the
Scala-Virtualized compiler. \code{JSLiteral} inherits from a special
trait, which indicates its \code{new} expressions should be
reified. So the \code{new} expression is turned into a method call,
with information about all the field definitions. A complication is
that a field definition might reference another field being defined
(such as \code{b} being defined in terms of \code{a} in the example
above). So each definition is represented by its name and a {\it
  function} which takes a self type for the object literal. These
definition functions are evaluated in an order which allows the self
references to be resolved.

@@

There are a few things worth noting here. First, the return type of
\code{foo} method is rewritten to be
\code{ControlContext}.\footnote{The \code{ControlContext} class
  implements the continuation monad and is provided by Scala's
  standard library. \code{ControlContext[A,B,C]} is similar to C\#'s
  \code{Task<T>}, but more general. In our discussion, we omit the
  type parameters for simplicity.} This is due to the fact that
\code{sleep} is used in the body of \code{foo}. Also, note that the
code to be executed after sleeping is captured as a continuation
(anynomous function) and passed to the \code{ControlContext} through a
call of the \code{map} method. The body of the \code{reset} block is
rewritten in a similar vein.

@@

\subsection{CPS for Suspendable Traversals}
We need to be able to suspend our execution while traversing an array in order to implement functionality from listing~\ref{code:twitter_example}. Let's consider a simplified example where we want to iterate over an array and sleep during each iteration. We present both code written in JavaScript and our DSL that achieves that (listings~\ref{code:js:sleep_iter}~\&~\ref{code:sleep_iter}).
\begin{figure}
\lstset{basicstyle=\ttfamily\scriptsize}
\begin{changemargin}{-0.7cm}{-0.7cm}
\setlength{\columnseprule}{0.5pt}
\begin{multicols}{2}
\begin{lstlisting}[caption=JavaScript,label=code:js:sleep_iter]
var xs = [1, 2, 3]
var i = 0
var msg = null
function f1() {
  if (i < xs.length) {
    window.setTimeout(f2, xs[i]*1000)
    msg = xs[i]
    i++
  } else {
    console.log("done")
  }
}
function f2() {
  console.log(msg)
  f1()
}
f1()
\end{lstlisting}
\begin{lstlisting}[escapechar=|,caption=DSL,label=code:sleep_iter]
val xs = array(1, 2, 3)
// shorthand for xs.suspendable.foreach
for (x <- xs.suspendable) {
  sleep(x * 1000)
  console.log(String.valueOf(x))
}
log("done")

@@

In the DSL code, we use a \code{suspendable} variant of arrays, which is achieved through an implicit conversion from regular arrays:
\begin{lstlisting}
implicit def richArray(xs: Rep[Array[A]]) = new {
  def suspendable: SArrayOps[A] = new SArrayOps[A](xs)
}
\end{lstlisting}
The idea behind \code{suspendable} arrays is that iteration over them can be suspended. We'll have a closer look at how to achieve that with the help of CPS. The \code{suspendable} method returns a new instance of the \code{SArrayOps} class defined here:

\begin{lstlisting}[caption=Suspendable foreach,label=code:suspendable_foreach]
class SArrayOps(xs: Rep[Array[A]]) {
  def foreach(yld: Rep[A] => Rep[Unit] @suspendable):
    Rep[Unit] @suspendable = {
      var i = 0
      suspendableWhile(i < xs.length) { yld(xs(i)); i += 1 }
    }
}
\end{lstlisting}
Note that one cannot use while loops in CPS but we can simulate them
with recursive functions. Let's see how a regular while loop can be
simulated with a recursive function with call-by-name parameters:
\begin{lstlisting}
def recursiveWhile(cond: => Boolean)(body: => Unit): Unit = {
  def rec = () => if (cond) { body; rec() } else {}
  rec()
}
\end{lstlisting}

@@

With the abstractions for suspendable loops and traversals at hand,
what remains to complete the Twitter example from the beginning of the section
is the actual Ajax request/response cycle.
